---
title: "Preparing the full date table"
#output: html_notebook
author: "Val Syverson"
---

This script assembles the tables "geochrons" and "alldates". The latter is the one called by most subsequent scripts.

```{r}
if ("here" %in% installed.packages()) {require(here)} else {install.packages("here"); require(here)}
i_am("final_scripts/2_get_alldates.Rmd")
source(here("final_scripts/functions/2_get_all_dates_functions.R"))
alldates <- read.csv(here("data/dates_for_analysis_smallmamm.csv"))
# if this does not exist, run 1_create_datetable.Rmd
```

## assign analysis unit names
```{r}
analunitnames <- data.frame(auid = unique(alldates$analysisunitid),auname = NA)
I <- 1
for (i in I:nrow(analunitnames)) {
  res <- try(get_from_tilia(values = analunitnames$auid[i], 
                            params = "analunitid", meth = "getanalysisunitbyid"))
  if (length(res$data) > 0) analunitnames$auname[i] <- res$data$analysisunitname
}
alldates$analysisunitname <- analunitnames[match(alldates$analysisunitid,analunitnames$auid),]$auname
```

## assign fake lab numbers to dates without lab numbers
```{r}
fakelabnumbers <- paste0("TEMP",1:length(which(is.na(alldates$labnumber))))
alldates[which(is.na(alldates$labnumber)),"labnumber"] <- fakelabnumbers
```


## flag human remains and rejected dates
```{r}
alldates$human <- FALSE
human_strings <- c("Teeth, Homo","Bone, Homo","human bone","Homo sapiens bone","Human skull")
alldates[which(apply(sapply(human_strings,function(x) grepl(x,alldates$materialdated)),1,any)),"human"] <- TRUE
alldates[which(apply(sapply(human_strings,function(x) grepl(x,alldates$notes)),1,any)),"human"] <- TRUE

rejected_strings <- c("Date rejected","date rejected"," too ","contaminated")
unique(alldates[which(apply(sapply(rejected_strings,function(x) grepl(x,alldates$notes)),1,any)),"notes"])
alldates[which(apply(sapply(rejected_strings,function(x) grepl(x,alldates$notes)),1,any)),"rejected"] <- TRUE
```

## category tagging
assign category names for each collection
- 1 = one analysis unit, uncalibrated 14C dates only
- 2 = multiple analysis units, uncalibrated 14C dates only
- 3 = analysis units with both uncalibrated 14C and other dates
- 4 = only non-14C or calibrated 14C dates (new chronologies will not be computed)
```{r}
write.csv(alldates,here("data/rawdates.csv"))
alldates$category <- NA
alldates.clean <- subset(alldates,alldates$infinite == FALSE & 
                                  alldates$rejected == FALSE & 
                                  alldates$human == FALSE)
collids.cat <- data.frame(collid = unique(alldates.clean$collectionunitid))
collids.cat$category <- NA
for (i in 1:length(collids.cat$collid)) {
  dd <- which(alldates.clean$collectionunitid == collids.cat$collid[i])
  if (any(alldates.clean[dd,]$agetypeid != "Radiocarbon years BP")) {
    if (any(alldates.clean[dd,]$agetypeid == "Radiocarbon years BP")) {
      collids.cat[i,"category"] <- 3 } else { collids.cat[i,"category"] <- 4 }
  } else {
    if (length(unique(alldates.clean[dd,]$analysisunitid)) == 1) {
      collids.cat[i,"category"] <- 1 } else { collids.cat[i,"category"] <- 2 }
  }
}
alldates$category <- collids.cat$category[match(alldates$collectionunitid,collids.cat$collid)]
write.csv(alldates,here("data/alldates_allcategories.csv"),row.names = F)
alldates <- subset(alldates,alldates$category != 4)
```

# save and delete extra items
```{r}
write.csv(alldates,here("data/alldates.csv"),row.names = F)
```

# clear workspace 
```{r}
#remove objects other than alldates
objectlist <- ls()
rm(list = objectlist[which(!objectlist == "alldates")]); rm(objectlist)
```

