---
title: "Generating oxcAAR chronologies"
#output: html_notebook
author: "Val Syverson"
---


# load functions
```{r}
if ("here" %in% installed.packages()) {require(here)} else {install.packages("here"); require(here)}
i_am("final_scripts/3_autogen_chronologies.Rmd")
source("functions/3_autogen_chronologies_functions.R")
```

# load dates table and split by category
```{r}
dates <- read.csv(here("data/alldates.csv"))
dates <- split(dates, dates$category)
```


# generate simple chronologies from category 1 analysis units

##grab the single analysis unit dates and get the ids, sorted from shortest to longest
```{r}
simple <- dates$`1`
simplecollids <- as.numeric(names(sort(table(simple$collectionunitid))))
```

## remove the collections that already have results in the sample ages folder
```{r}
done <- as.numeric(unlist(sapply(dir(here("outputs/sampleages/")),
                  function(x) return(strsplit(x,".csv")))))
simplecollids <- subset(simplecollids, !simplecollids %in% done)
```

## generate the results
```{r}
J <- 1
for (j in J:length(simplecollids)) {
  print(paste0(c("Collection",j,"of",length(simplecollids)),collapse=" "))
	result <- dateranges(simplecollids[j],data = simple)
 	write.csv(result$sampleages,
 	          paste0(here("outputs/sampleages/"), simplecollids[j],".csv", collapse = ""), row.names = F)
 	write.csv(result$chroncontrols,
 	          paste0(here("outputs/chroncontrols/"),simplecollids[j],".csv",collapse = ""),row.names = F)
}
```

# Multi-analysis-unit collections

#Get the data
```{r}
multi <- dates$`2`
multi <- split(multi, multi$collectionunitid)
```

Remove the ones that have already been run.
```{r}
done <- as.numeric(unlist(sapply(dir(here("outputs/sampleages")),
                                 function(x) return(strsplit(x,".csv")))))
multi <- multi[!names(multi) %in% done]
```

## tagging ordered collections (manual)
Some of the collections have analysis units whose names indicate that they're ordered. In these cases, we're going to add a column called "order" that tells the OxCal assembly function to put them in the right order and wrap them in a Sequence(), which can increase the precision of the estimate. Unfortunately this has to be done one by one, manually.

```{r}
i <- 0
```

We page through the sets of analysis unit names one by one...
```{r}
i <- i + 1
temp_order <- data.frame(
  meanage = rev(sort(by(multi[[i]]$age,multi[[i]]$analysisunitname,mean,na.rm = T)))
  );temp_order
```

Label them as ordered /only/ if the analysis unit names imply a stratigraphic order /and/ they correspond more-or-less to their actual mean ages so you can get the direction right. (Don't rely on the auto-generated order here; go in and change them manually! There WILL be mistakes.) 

If they aren't ordered, /don't run this block/.
```{r}
temp_order$order <- 1:nrow(temp_order)
temp_order <- edit(temp_order)
multi[[i]]$auorder <- temp_order[match(multi[[i]]$analysisunitname, rownames(temp_order)),"order"]
```

Go on to the next one and repeat.

## generating chronologies

Compute the chronologies for your collections.
```{r}
J <- 1
for (j in J:length(multi)) {
  print(paste0(c("Collection",j,"of",length(multi)),collapse=" "))
  result <- dateranges.multi(multi[j])
  write.csv(result$sampleages,
            paste0(here("outputs/sampleages/"),names(multi)[j],".csv",collapse = ""),row.names = F)
  write.csv(result$chroncontrols,
            paste0(here("outputs/chroncontrols/"),names(multi)[j],".csv",collapse = ""),row.names = F)
  }
```

# clear workspace 
```{r}
#remove objects other than dates
objectlist <- ls()
rm(list = objectlist[which(!objectlist == "dates")]); rm(objectlist)
```