---
title: "Single site"
author: "Val Syverson
---
12 Nov 2024

This script is to be used to generate chronologies according to the same method used for the Syverson and Blois (2025) chronologies, for example, if the geochronologic dataset for a collection has been updated in Neotoma.

* The first part of this script acquires all dates in Neotoma for collections from a single site or set of sites, specified by siteid. 
* The section "Run chronologies" runs the OxCal analysis for the dates in the sites you specify and saves the results (sample ages and chron controls) over any existing chronologies.
* The section "Compile chronologies" creates tables that can be inserted into the "chronologies", "sampleages", and "chroncontrols" tables in the database.
* The section "Add to main date table" adds any new dates to the main date table (data/alldates.csv). This part of the script is only necessary to run if you are planning to reanalyze the whole table, which is usually not going to be the case; thus, it is probably not useful, but is included for completeness.



# Download all dates for one or several specified Neotoma sites (by siteid)
## load libraries
```{r}
if ("here" %in% installed.packages()) {require(here)} else {install.packages("here"); require(here)}
i_am("final_scripts/6_get_single_site.Rmd")
source(here("final_scripts/functions/6_single_site_functions.R"))
```

## specify siteid to be downloaded 
```{r}
site_selection <- 3586
# this siteid is just an example; change this line to the siteid of the site that you want to download 
add.dates <- get_site_geochrons(site_selection,geochronology.table = geochronology.table,agelimit = 30000)
```

## or specify multiple siteids to be downloaded
```{r}
site_selections <- c(5120,5700,16549)
# alternatively, download multiple sites by listing the siteids
# again, these are example siteids; replace them with the ids of the desired sites
add.dates <- vector(mode = "list",length = length(site_selections)); names(add.dates) <- site_selections
for (i in 1:length(site_selections)){
  add.dates[[i]] <- try(get_site_geochrons(site_selections[i]))
}
add.dates <- list.stack(add.dates[sapply(add.dates,function(x) 
          class(x) == "data.frame" & !all(is.na(x)))])
no.dates <- subset(site_selections,!site_selections %in% unique(add.dates$siteid))
print(paste0(c("No dates from sites",no.dates),collapse=", "))
```

# Run chronologies
Re-run chronology scripts for all affected collections.

## load scripts 
```{r}
source(here("final_scripts/functions/3_autogen_chronologies_functions.R"))
```

## select only dates from affected collections
```{r}
dates <- split(add.dates,add.dates$category)
```

## compute chronologies by category
### single analysis unit collection
```{r}
if (!is.null(dates$`1`)) {
  simple <- dates$`1`; simplecollids <- unique(simple$collectionunitid)
  J <- 1
  for (j in J:length(simplecollids)) {
    print(paste0(c("Collection",j,"of",length(simplecollids)),collapse=" "))
  	result <- dateranges(simplecollids[j],data = simple)
   	write.csv(result$sampleages,
     paste0(here("outputs/sampleages/"), "/", simplecollids[j],".csv", collapse = ""), row.names = F)
   	write.csv(result$chroncontrols,
     paste0(here("outputs/chroncontrols/"), "/", simplecollids[j],".csv",collapse = ""),row.names = F)
  }
}
```

### multiple analysis unit collection
```{r}
if (!is.null(dates$`2`)){
  multi <- dates$`2`
  multi <- split(multi, multi$collectionunitid)
  J <- 1
  for (j in J:length(multi)) {
    print(paste0(c("Collection",j,"of",length(multi)),collapse=" "))
    result <- dateranges.multi(multi[j])
    write.csv(result$sampleages,
              paste0(here("outputs/sampleages/"), "/", names(multi)[j],".csv",collapse = ""),row.names = F)
    write.csv(result$chroncontrols,
              paste0(here("outputs/chroncontrols/"), "/", names(multi)[j],".csv",collapse = ""),row.names = F)
    }
}
```

### collection containing some non-radiocarbon dates
```{r}
if (!is.null(dates$`3`)){
  mixed <- dates$`3`
  mixed <- split(mixed,mixed$collectionunitid)
  source(here("final_scripts/functions/6_manual_chronologies_functions.R"))
  codeblocks <- sapply(mixed,makeOxcalCode)
  for (i in 1:length(codeblocks)){
    filename <- paste0(c(here("scripts/oxcalscripts/oxcode"),names(codeblocks[i]),".txt"),
                       collapse = "")
    print(filename)
    write(codeblocks[[i]],file = filename)
  }
  print(c(paste0("New OxCal scripts generated for collection units ", 
               paste0(sort(names(mixed)),collapse = ", "),
               ".",collapse=" "),
          "Run them in the OxCal web interface.",
          "Then ingest them using script 4_manual_chronologies.Rmd, starting at chunk 4."))
}
```

# Compile chronologies
## Load data from files created in previous section
```{r}
collids <- unique(add.dates$collectionunitid)
fpath <- here("outputs/sampleages")
names <- data.frame(collid = gsub(".csv","",dir(path = fpath)),
                    path = unname(sapply(dir(path = fpath),
                       function(x) paste(c(fpath,"/",x),collapse = ""))))
names <- subset(names,suppressWarnings(!is.na(as.numeric(names$collid))) & grepl("csv",names$path))
names <- names[sapply(collids,function(x) which(grepl(x,names$collid))),]
sampleages <- list.stack(apply(names,1,function(x) {
  csv <- try(read.csv(file = x["path"]),silent = T)
  if (class(csv) == "data.frame") {
      return(suppressWarnings(cbind(collid = as.numeric(x["collid"]),csv)))}
  }),use.names = F)
colnames(sampleages)[1] <- "collectionunitid"
inferred_sampleages <- read.csv(here("outputs/inferred_sampleages.csv"))[,1:6]
if (any(sampleages$collectionunitid %in% inferred_sampleages$collectionunitid)) {
  inferred_sampleages$source <- "inferred"
  inferred_sampleages <- subset(inferred_sampleages,
                                inferred_sampleages$collectionunitid %in% sampleages$collectionunitid)
  sampleages$source <- "computed"
  sampleages <- as.data.frame(rbind(sampleages,inferred_sampleages))
}
future <- which(sampleages$sampleages.ageyounger < -75)
sampleages[future,"sampleages.ageyounger"] <- -75
fpath2 <- here("outputs/chroncontrols")
names2 <- data.frame(collid = gsub(".csv","",dir(path = fpath2)),
                    path = unname(sapply(dir(path = fpath2),
                       function(x) paste(c(fpath2,"/",x),collapse = ""))))
names2 <- subset(names2,suppressWarnings(!is.na(as.numeric(names2$collid))))
names2 <- names2[sapply(collids,function(x) which(grepl(x,names2$collid))),]
chroncontrols <- list.stack(apply(names2,1,function(x) {
  csv <- try(read.csv(file = x["path"]),silent = T)
  if (class(csv) == "data.frame") {
      return(suppressWarnings(cbind(collid = as.numeric(x["collid"]),csv)))}
  }),use.names = F)
colnames(chroncontrols)[1] <- "collectionunitid"
```

## Assign types
```{r}
chroncontroltypes <- get_table("chroncontroltypes")[,1:3]
temp <- left_join(chroncontrols,alldates[,c("geochronid","geochrontypeid","agetypeid")])
table(temp$geochrontypeid)
typeindex <- data.frame(table(temp$geochrontypeid))
```


```{r}
#Each type of chroncontrol needs a label and there is no way to do this automatically
#If there are fewer or more types than listed here, this will throw an error

#typeindex$chroncontroltype <- c("Palaeomagnetic","Argon-argon",
#                                "Radiocarbon, calibrated, Bayesian modelled",
#                                "Optically stimulated luminescence",
#                                "Thermoluminescence","Uranium-series")

#Adjust this list as necessary so that each type of chroncontrol has a matching value in the chroncontroltypes table in the following join.
#For the example collection Crystal Ball Cave:
typeindex$chroncontroltype <- "Radiocarbon, calibrated, Bayesian modelled"

typeindex <- left_join(typeindex[c("Var1","chroncontroltype")],chroncontroltypes)
temp <- left_join(temp,typeindex[,c("Var1","chroncontroltype","chroncontroltypeid")],
                  by = join_by(geochrontypeid == Var1))

temp[which(temp$geochrontypeid == "Carbon-14" & temp$agetypeid %in% c("calendar yr BP","cal yr BP")),"chroncontroltype"] <- "Calibrated radiocarbon years BP"
temp[which(temp$geochrontypeid == "Carbon-14" & temp$agetypeid %in% c("calendar yr BP","cal yr BP")),"chroncontroltypeid"] <- 17
chroncontrols <- temp[,c(colnames(chroncontrols),"chroncontroltypeid")]
```

## Generate table of ranges and temporary chronids
```{r}
collchrons <- list.stack(tapply(sampleages,sampleages$collectionunitid,
                      dfbounds, simplify = F))
timeinfo <- cbind(filename = dir(here("outputs/sampleages")),
                  list.stack(sapply(dir(here("outputs/sampleages"),full.names = T), 
                                    file.info, simplify = F)))
timeinfo$collectionunitid <- unlist(sapply(timeinfo$filename,function(x) strsplit(x,".csv")))
timeinfo$dateprepared <- unlist(sapply(timeinfo$mtime,function(x) strsplit(as.character(x)," ")[[1]][1]))
collchrons <- merge(collchrons,timeinfo[,c("collectionunitid","dateprepared")])
collchrons$chronologies.notes <- NA
collchrons$TEMP_chronid <- paste0("TEMP",1:nrow(collchrons))
sampleages <- left_join(sampleages,                          
                        collchrons[,c("collectionunitid","agemodel","TEMP_chronid")],
                          by = join_by(collectionunitid == collectionunitid,
                                       chronology.agemodel == agemodel), 
                          relationship = "many-to-one")
alldates <- read.csv(here("data/alldates.csv"))
temp <- distinct(inner_join(chroncontrols,alldates[,c("geochronid","analysisunitid")]))
temp[,c(1:2,4:8)] <- apply(temp[,c(1:2,4:8)],2,as.numeric)
sampleages[,c(1:2,4:6)] <- apply(sampleages[,c(1:2,4:6)],2,as.numeric)
newchron.sampleages <- split(sampleages,sampleages$chronology.agemodel)
templist <- list(bounds = distinct(left_join(temp,
                            distinct(newchron.sampleages$bounds[,c("analysisunitid","TEMP_chronid")]))), 
                 event = distinct(left_join(temp,
                            distinct(newchron.sampleages$event[,c("analysisunitid","TEMP_chronid")]))))
templist$bounds <- left_join(templist$bounds,subset(sampleages,sampleages$chronology.agemodel == "bounds")[,c("analysisunitid","TEMP_chronid")],relationship = "many-to-many")
templist$event <- left_join(templist$event,subset(sampleages,sampleages$chronology.agemodel == "event")[,c("analysisunitid","TEMP_chronid")],relationship = "many-to-many")
chroncontrols <- distinct(list.stack(templist))
inferred_chroncontrols <- read.csv(here("outputs/inferred_chroncontrols.csv"))
inferred_chroncontrols <- distinct(subset(inferred_chroncontrols, inferred_chroncontrols$collectionunitid %in% collchrons$collectionunitid))
temp <- distinct(left_join(inferred_chroncontrols[,1:9],sampleages[,c("collectionunitid","chronology.agemodel","TEMP_chronid")],by = join_by(collectionunitid,chronology.agemodel), relationship = "many-to-many"))
chroncontrols <- distinct(rbind(chroncontrols,temp[,-9]))
future_chronids <- unique(sampleages[future,"TEMP_chronid"])
collchrons[which(collchrons$TEMP_chronid %in% future_chronids),
   "chronologies.notes"] <- "younger age limits truncated at publication date (2025 CE)"
```

## Save new tables
```{r}
filestring <- paste0(here("outputs/"),paste0(collids,collapse = "_"),collapse = "")
write.csv(chroncontrols,paste0(c(filestring,"_caldates.csv"),collapse = ""), row.names = F)
write.csv(collchrons,paste0(c(filestring,"_chronologies.csv"),collapse = ""), row.names = F)
write.csv(sampleages,paste0(c(filestring,"_sampleages.csv"),collapse = ""), row.names = F)
```


# Add to main date table
## search "alldates" table for duplicates of existing dates
```{r}
alldates <- read.csv(here("data/alldates.csv"))
nas <- which(is.na(add.dates$labnumber)) #if there are NAs
maxtemp <- max(as.numeric(sapply(grep("(^TEMP)",alldates$labnumber,value = T),substr,5,1000)))
add.dates[nas,"labnumber"] <- paste("TEMP",(maxtemp+1):(maxtemp+length(nas)),sep="")
check <- which(add.dates$labnumber %in% alldates$labnumber)
print(check) #should be zero, otherwise check
if (length(check) == 0) {changed <- check}
```

## check table for any changes to dates that were already included in the table
```{r}
check.rev <- na.omit(match(add.dates[check,]$labnumber,alldates$labnumber))
alldates[which(!apply(alldates[check.rev,c("age","errorolder","erroryounger")] == 
  add.dates[check,c("age","errorolder","erroryounger")],1,all)),"labnumber"] -> changed
length(changed)
```

## drop duplicates and add only the new dates ("add.dates")
```{r}
alldates <- rbind(subset(alldates,!alldates$labnumber %in% changed),
  subset(add.dates,add.dates$labnumber %in% changed))
add.dates <- subset(add.dates,!add.dates$labnumber %in% changed 
                    & !add.dates$labnumber %in% alldates$labnumber)
check <- which(add.dates$labnumber %in% alldates$labnumber)
print(check) #should be zero, otherwise check
```

## update saved copy of "alldates" table with new dates
```{r}
if (length(check) == 0) { 
  alldates <- distinct(rbind(alldates, add.dates[,colnames(alldates)]))
  write.csv(alldates,here("data/alldates.csv"),row.names = F)
  print(paste("Added",nrow(add.dates),"dates, changed",length(changed),"dates, and saved to data/alldates.csv"))
} else {
  print(subset(alldates,alldates$labnumber %in% add.dates$labnumber))
  print(add.dates)
}
```
