```{r load packages}
if ("here" %in% installed.packages()) {require(here)} else {install.packages("here"); require(here)}
i_am("final_scripts/7_make_pub_figures.Rmd")
source(here("final_scripts/functions/7_figures_functions.R"))
required.packages <- c("leaflet","maps","naturalearth","sf","terra","ggplot2","dplyr","neotoma2","reshape","rnaturalearth","viridis","jsonlite","httr","rlist")
checkAndInstall(required.packages)
```

```{r refresh data}
rm(alldates,geochrons,newchron.range,newchron.all,found,corrected)
```


```{r get data}
if (!exists("alldates")) alldates <- read.csv(here("data/alldates.csv"))
if (!exists("geochrons")) geochrons <- read.csv(here("data/dates_for_analysis_smallmamm.csv"))
if (!exists("newchron.range")) newchron.range <- read.csv(here("outputs/new_chronologies.csv"))
if (!exists("newchron.all")) newchron.all <- read.csv(here("outputs/new_sampleages.csv"))
if (!exists("newchron.ctrls")) newchron.ctrls <- read.csv(here("outputs/new_caldates.csv"))
if (!exists("found")) found <- read.csv(here("data/found.csv"))
if (!exists("corrected")) {
  corrected <- read.csv(here("data/corrections.csv"))
  corrected <- data.frame(siteid = corrected$Site.ID, sampleid = corrected$Sample.ID,
                          geochronid = corrected$Geochron.ID, labnumber = corrected$Lab..)
  corrected[which(corrected$labnumber == ""),"labnumber"] <- NA
  corrected <- distinct(subset(corrected, !is.na(corrected$geochronid) | !is.na(corrected$labnumber)))
}
```

```{r tag dates by update status}
alldates$source <- "Neotoma"
alldates[which(alldates$labnumber %in% found$ndb.geochronology.labnumber &
                 !is.na(alldates$labnumber)),"source"] <- "Corrected"
alldates[which(alldates$geochronid > 51354),"source"] <- "Added"
```

```{r get site information}
siteids <- unique(alldates$siteid)
sitelist <- get_sites(siteids, all_data = T)
```

```{r parse site info into dataframe}
sites <- as.data.frame(sitelist)
sites <- inner_join(sites,getids(sitelist))
sites <- distinct(sites[,-which(colnames(sites) == "datasetid")])
sites <- subset(sites,sites$collunitid %in% newchron.range$collectionunitid)
colnames(sites)[9] <- "collectionunitid"
sites[,c("siteid","collectionunitid")] <- apply(sites[,c("siteid","collectionunitid")],2,as.numeric)
```

```{r tag sites by update status}
sites$newdates <- "some"
sites[which(!sites$siteid %in% subset(alldates, 
          alldates$source == "Added" | alldates$source == "Updated")$siteid),"newdates"] <- "none"
sites[which(!sites$siteid %in% subset(alldates, 
          alldates$source == "Neotoma")$siteid),"newdates"] <- "all"
sites$newdates <- factor(sites$newdates,levels = c("none","some","all"))
sites <- sites[order(sites$newdates),]
```


```{r add collection age ranges}
#note that this doubles the size of the table (two chronologies per collection)
site_chrons <- distinct(inner_join(sites,
                    newchron.range[,c("collectionunitid","ageboundolder","ageboundyounger","agemodel")]))
```


```{r Figure 2a objects}
epsg9311 <- leafletCRS(
  crsClass = "L.Proj.CRS",
  code = "EPSG:9311",
  proj4def = "+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs",
  resolutions = 2 ^ (16:7),
  origin = c(0,0)
)

north_america <-
  rnaturalearth::countries110 |>
  dplyr::filter(CONTINENT == "North America")

pal <- colorFactor("RdYlBu",sites$newdates,reverse = T)
```


```{r Figure 2a}
figure2a <- leaflet(north_america, options = leafletOptions(crs = epsg9311)) %>%
  addPolygons(weight = 1,color = "grey70bl",opacity = 0.8) %>%
  addCircleMarkers(data = sites[,c("long","lat")],
                   color = ~pal(sites$newdates),fillColor = ~pal(sites$newdates),
                   radius = 2,stroke = T,fillOpacity = 0.9) %>%
  addLegend(position = "bottomleft", pal = pal, values = levels(sites$newdates),
            title = "New dates by site")
```


```{r Figure 2b objects}
crdref <- "+proj=longlat +datum=WGS84"
sites.vect <- vect(sites[c("long","lat")],geom = c("long","lat"),crs = crdref)
dates <- inner_join(distinct(alldates[,c("collectionunitid","geochronid","age","errorolder")]),
                    distinct(sites[,c("collectionunitid","siteid","sitename","lat","long","newdates")]))
dates$interval <- cut(as.numeric(dates$age),c(0,6000,11700,16000,60000))
levels(dates$interval) <- c("Late Hol.","Early Hol.","Deglacial","LGM")
dates$new <- as.factor(dates$newdates)
levels(dates$new) <- c("FALSE","TRUE","TRUE"); dates$new <- as.factor(as.character(dates$new))
temp <- subset(dates,dates$new == TRUE)
dates.vect <- sapply(split(temp,temp$interval), 
                     function(x) vect(x[,c("long","lat")],geom = c("long","lat"),crs = crdref))
```


```{r Figure 2b shapefile}
if (!file.exists(here("data/na_cec_eco_l1/NA_CEC_Eco_Level1.shp"))) {
  if (! file.exists(here("data/na_cec_eco_l1.zip"))) {
    print("Download Level 1 Ecoregions Shapefile from epa.gov/eco-research/ecoregions-north-america into data folder")
    } else {
  unzip(here("data/na_cec_eco_l1.zip"), exdir = here("data/na_cec_eco_l1"))}}
eco1 <- read_sf(here("data/na_cec_eco_l1/NA_CEC_Eco_Level1.shp"))
eco1 <- st_transform(eco1,st_crs("+proj=longlat +datum=WGS84"))
temp <- sapply(dates.vect,function(x) relate(vect(eco1), x, "contains"))
datecount <- sapply(temp, function(x) apply(x,1,sum))
```

```{r print Figure 2b}
temp <- data.frame(eco1$NA_L1NAME,datecount)
colnames(temp) <- c("Ecoregion","LH","EH","DG","LGM")
temp <- do.call("rbind",by(temp,temp$Ecoregion,function(x) apply(x[,2:5],2,function(x) sum(x))))
datecounts <- data.frame(Ecoregion = rownames(temp), temp[,4:1])
datecounts$Ecoregion <- gsub("NORTHWESTERN","NW",datecounts$Ecoregion)
datecounts$Ecoregion <- reorder(datecounts$Ecoregion,
                                apply(datecounts[,2:5],1,sum),decreasing = T)
datecounts <- subset(datecounts,apply(datecounts[,2:5],1,sum) != 0)
datecounts <- datecounts[order(apply(datecounts[,2:5],1,sum),decreasing = T),]
temp <- reshape::melt(datecounts);colnames(temp) <- c("Ecoregion","Interval","Count")
figure2b <- ggplot(temp,aes(x = Interval,y = Count,fill = Ecoregion)) + geom_bar(stat = "identity") + scale_fill_brewer(palette = "Paired")
```

```{r figure 5 objects}
newchron.event <- subset(newchron.all,newchron.all$chronology.agemodel == "event")
newchron.event$agediff_new <- newchron.event$sampleages.ageolder - newchron.event$sampleages.ageyounger
newchron.event$rangebin <- cut(newchron.event$agediff_new,
                               breaks = quantile(newchron.event$agediff_new,probs = seq(0,1,0.05)))
newchron.event$age <- apply(newchron.event,1,function(x) {
  if (is.na(x["sampleages.age"])) {
    return(mean(as.numeric(c(x["sampleages.ageolder"],x["sampleages.ageyounger"]))))
  } else return(as.numeric(x["sampleages.age"]))
})
newchron.event$agebin <- cut(log(newchron.event$age),
                             breaks = quantile(log(newchron.event$age),probs = seq(0,1,0.05)))
levels(newchron.event$agebin) <- rev(viridis::viridis(n = 20))

temp <- dplyr::inner_join(sites,newchron.event[,c("age","collectionunitid","analysisunitid","agebin","agediff_new","rangebin")],relationship = "many-to-many")
newchron_sites_event <- dplyr::distinct(temp[,c("age","agebin","agediff_new","rangebin","lat","long")])
newchron_sites_event <- subset(newchron_sites_event,newchron_sites_event$agediff_new < 10000)
newchron_sites_event <- newchron_sites_event[rev(order(newchron_sites_event$agediff_new)),]
newchron_sites_event$range_scaled <- (newchron_sites_event$agediff_new/max(newchron_sites_event$agediff_new))
```


```{r generate figure 5}
figure5 <- leaflet(north_america, options = leafletOptions(crs = epsg9311)) %>%
  addPolygons(weight = 1, color = "grey70", opacity = 0.8) %>%
  addCircleMarkers(data = newchron_sites_event, opacity = 0.2,
     radius = 15*newchron_sites_event$range_scaled,
     color = newchron_sites_event$agebin, group = "circles") %>%
  addLegend(position = "bottomleft",
      pal = colorNumeric(palette = "viridis", newchron_sites_event$age, reverse = T), 
      values = ~newchron_sites_event$age, title = "Age (ybp)")
figure5
```


```{r Figure 6b}
#range plot of age bounds
if (!exists("newchron.all")) newchron.all <- read.csv(here("outputs/new_sampleages.csv"))
test <- subset(newchron.all,newchron.all$chronology.agemodel == "bounds")[,c("sampleages.ageolder","sampleages.ageyounger")]
test <- test[order(test$sampleages.ageolder),]
#test <- test[-which(test$sampleages.ageolder == max(test$sampleages.ageolder)),] #drop outlier site
test <- test[-which(test$sampleages.ageolder > 150000),]
Figure6b <- ggplot(test, aes(y = 1:nrow(test))) +
  geom_linerange(aes(xmin = sampleages.ageyounger, xmax = sampleages.ageolder),
                 linewidth = 1.5, alpha = 0.25) +
  scale_x_reverse() +
  geom_point(aes(x = sampleages.ageolder), color = "blue", size = 0.5) +
  geom_point(aes(x = sampleages.ageyounger), color = "red", size = 0.5) +
  geom_vline(xintercept = 11700) +
  ylab("Samples ordered by maximum age") + xlab("Sample age range (cal yr bp)")
```
# Statistics
## Download existing default chronologies to be replaced
```{r download chronology ranges}
chrons <- vector(mode = "list",length = length(unique(newchron.all$collectionunitid)))
i <- 0
I <- i + 1; J <- 100
while (I <= length(chrons)) {
  for (i in I:J) {
    x <- unique(newchron.all$collectionunitid)[i]
    temp <- get_from_tilia(meth = "getchronologiesbycollunitid", 
                                param = "collectionunitid", 
                                value = x)
    if (length(temp$data) > 0) {
                   chrons[[i]] <- data.frame(collectionunitid = as.numeric(x),temp$data)}
  }
  I <- i + 1; J <- min(c(length(chrons),(i + 100)))
}
chrons <- list.stack(chrons)
chrons <- distinct(subset(chrons,chrons$isdefault == TRUE)[,c("collectionunitid","chronologyid","chronologyname","ageboundolder","ageboundyounger","agetype","agemodel")])
write.csv(chrons,here("data/old_chrons.csv"))
```

## Combine old and new collection age ranges
```{r join with newchron.range}
compare_chrons <- right_join(chrons,newchron.range,by = "collectionunitid",relationship = "many-to-many")[,1:10]
colnames(compare_chrons) <-  c("collectionunitid","old.chronid","old.chronname","old.ageboundolder","old.ageboundyounger", "old.agetype","old.agemodel","new.agemodel","new.ageboundolder","new.ageboundyounger")
compare_chrons <- split(compare_chrons,compare_chrons$new.agemodel)
```

## Download existing analysis unit (sample) age ranges
```{r download sample ages}
sampleages <- vector(mode = "list",length = length(unique(chrons$chronologyid)))
i <- 0
I <- i + 1; J <- i + 100
while (I <= length(sampleages)) {
  for (i in I:J) {
    x <- unique(chrons$chronologyid)[i]
    temp <- get_from_tilia(meth = "getsampleagesbychronid",
                           param = "chronologyid",
                           value = x)
    if (length(temp$data) > 0){
      sampleages[[i]] <- data.frame(chronologyid = as.numeric(x),temp$data)
    }
  }
  I <- i + 1; J <- min(c(nrow(sampleages),(i + 100)))
}
sampleages.df <- distinct(list.stack(sampleages, fill = TRUE))[,1:5]
```

## Join with downloaded chronologies
```{r download sample ages}
oldchron_samp <- data.frame(dplyr::inner_join(chrons, sampleages.df, by = "chronologyid"))
```
## Download sample ids and match with analysis unit ids in new sample ages
```{r download sample ids}
sampleids <- vector(mode = "list",length = length(unique(oldchron_samp$sampleid)))
i <- 0
I <- i + 1; J <- 100
while (I <= length(sampleids)) {
  for (i in I:J) {
    x <- unique(oldchron_samp$sampleid)[i]
    temp <- get_from_tilia(meth = "getsampleparents",
                           param = "sampleid",
                           value = x)
    if (length(temp$data) > 0){
      sampleids[[i]] <- data.frame(sampleid = as.numeric(x),analysisunitid = temp$data$analysisunitid)
    }
  }
  I <- i + 1; J <- min(c(nrow(sampleids),(i + 100)))
}
sampleids.df <- distinct(list.stack(sampleids)[,c("sampleid","analysisunitid")])
```

## Join sample ages to chronologies
```{r download sample ages}
oldsampleages <- inner_join(oldchron_samp,sampleids.df,by = "sampleid")
compare_sampleages <- left_join(distinct(newchron.all[,1:6]),distinct(oldsampleages),
                             by = c("analysisunitid","collectionunitid"), relationship = "many-to-many")
colnames(compare_sampleages) <- c("collectionunitid","analysisunitid","new.chrontype","new.age", "ageolder_new","ageyounger_new","old.chronologyid","old.chronologyname","old.chron.ageolder","old.chron.ageyounger","old.agetype","old.agemodel","sampleid","old.sample.age","old.sample.ageyounger","old.sample.ageolder")
compare_sampleages <- split(compare_sampleages,compare_sampleages$new.chrontype)
```

## Calculate differences
```{r calculate differences between old and new sample ages & chronology ranges}
#chronology ranges
compare_chrons <- sapply(compare_chrons, function(x) {
  old.agediff <- x$old.ageboundolder - x$old.ageboundyounger
  new.agediff <- x$new.ageboundolder - x$new.ageboundyounger
  ageshift_older <- x$old.ageboundolder - x$new.ageboundolder
  ageshift_younger <- x$old.ageboundyounger - x$new.ageboundyounger
  return(data.frame(x, old.agediff, new.agediff, ageshift_older, ageshift_younger))
}, simplify = F)

#sample ranges
compare_sampleages <- sapply(compare_sampleages, function(x) {
  old.agediff <- x$old.sample.ageolder - x$old.sample.ageyounger
  new.agediff <- x$ageolder_new - x$ageyounger_new
  ageshift_older <- x$old.sample.ageolder - x$ageolder_new
  ageshift_median <- x$old.sample.age - x$new.age
  ageshift_younger <- x$old.sample.ageyounger - x$ageyounger_new
  return(data.frame(x, old.agediff, new.agediff, ageshift_older, ageshift_median, ageshift_younger))
}, simplify = F)
```

## Report values in Table 2/Figure 6a
```{r}
Table2 <- cbind(
  t(sapply(rev(compare_sampleages), function(x) {
  return(data.frame(Sample_Age = -median(x$ageshift_median, na.rm = T),
                    Sample_Older = -median(x$ageshift_older, na.rm = T),
                    Sample_Younger = -median(x$ageshift_younger, na.rm = T)))
  })),
  t(sapply(rev(compare_chrons), function(x) {
  return(data.frame(Chron_Older = -median(x$ageshift_older, na.rm = T),
                    Chron_Younger = -median(x$ageshift_younger, na.rm = T)))
  }))
)

```

## Report values in "Computing new age estimates"
```{r}
rawdates <- read.csv(here("data/rawdates.csv"))
cleandates_report(rawdates)
#number of bad dates in various categories

allcategories <- read.csv(here("data/alldates_allcategories.csv"))
length(unique(subset(allcategories,allcategories$category == 4)$collectionunitid))
#number of collections with no radiocarbon dates

clean_dates <- cleandates(alldates)
data.frame(Carbon14 = length(which(clean_dates$geochrontypeid == "Carbon-14")),
                             Other = length(which(clean_dates$geochrontypeid != "Carbon-14")))
#date type counts
apply(clean_dates[,c("siteid","collectionunitid","analysisunitid","geochronid")],2,
      function(x) length(unique(x)))
#counts of unique ids

table(clean_dates$agetypeid)
table(clean_dates$geochrontypeid)
```

## Technical validation section
```{r dates}
newdates <- read.csv(here("outputs/newdates-uploaded.csv"))
apply(newdates[,c("ndb.geochronology.geochronid","ndb.collectionunits.collectionunitid","ndb.sites.siteid")],2,function(x) length(unique(x)))
```

```{r chronologies}
length(unique(newchron.range$collectionunitid))
apply(newchron.all,2,function(x) length(unique(x)))
test <- subset(newchron.all,newchron.all$chronology.agemodel == "event")
table(test$source)
test <- subset(newchron.range,newchron.range$agemodel == "event")
length(which(!test$collectionunitid %in% chrons$collectionunitid))
length(which(test$collectionunitid %in% chrons$collectionunitid))
dim(chrons)
table(chrons$agetype)
sapply(compare_sampleages, function(x) {
  return(data.frame(
    old.agediff = median(x$old.agediff, na.rm=T),
    new.agediff = median(subset(x$new.agediff, !is.na(x$old.agediff)), na.rm=T),
    difference = median(x$new.agediff - x$old.agediff, na.rm=T))
  )
})
sapply(compare_chrons, function(x) {
  return(data.frame(
    old.agediff = median(x$old.agediff, na.rm=T),
    new.agediff = median(subset(x$new.agediff, !is.na(x$old.agediff)), na.rm=T),
    difference = median(x$new.agediff - x$old.agediff, na.rm=T))
  )
})
```

